---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: Erin Ansbro
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 getting set up
getwd()
library(tidyverse)
library(agricolae)

Litter <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",stringsAsFactors = TRUE)

Litter$sampledate <- as.Date(Litter$sampledate , format = "%m/%d/%y")


#2 setting a theme
mytheme <- theme_classic(base_size = 13) +
  theme(axis.text = element_text(color = "navy"), 
        legend.position = "bottom")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes
Ha: Mean lake temperature recorded during July does change with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 scatter plot temp by depth
Litter <- mutate(Litter, Month=month(sampledate))

Litter$Month <-factor(  
  Litter$Month, levels = c(1,2,3,4,5,6,7,8,9,10,11,12),labels =c(  
    "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep",  
    "Oct", "Nov", "Dec") )

LitterCleaned <- Litter %>%
  filter(Month=="Jul")%>%
  select(lakename, year4, daynum, depth, temperature_C)%>%
  na.omit()

LitterScatter <- ggplot(LitterCleaned, aes(x=depth, y=temperature_C))+
  geom_point()+
  geom_smooth(method="lm", col="blue")+
  ylim(0,35)+
  labs(y="Temperature (C)", x="Depth (m)")
  
print(LitterScatter)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As depth increases, temperature decreases. From the plot, from 0 to 2 meters the temperature is consistently about 15 degrees Celsius. Once you hit over 2 meters it drops fairly linear to about 5 meters where the temperaure levels off to 5 degrees Celsius from there on out. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 linear regression for temp by depth
LitterLM <-lm(data = LitterCleaned, temperature_C~depth)
summary(LitterLM)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The null hypothesis should be rejected because the p-value is < 2.2e-16. This means, the alternative hypothesis that mean lake temperature recorded during July does change with depth across all lakes has statistical signifance. The R-squared is .7387, meaning change in depth account for about 74% of the reason why temperature changes. The degrees of freedom is 9726. Temperature is predicted to decrease by 1.94621 degrees when depth increases by 1 meter.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 AIC for explanatory variables
LitterAIC <- lm(data=LitterCleaned, temperature_C~year4+daynum+depth)

step(LitterAIC)

#10 lm based on AIC
LitterModel <- lm(data=LitterCleaned, temperature_C~year4+daynum+depth)
summary(LitterModel)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables are year4, daynum, and depth. This variance explains a little over 74% of temperature change. It is an improvement, but barely, at only a .24% increase from the 73.87% variance that was explained by depth only. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 ANOVA for lakes
summary(LitterCleaned$lakename)

LitterLakeName <- aov(data=LitterCleaned, temperature_C~lakename)
summary(LitterLakeName)

LitterLakeName2 <- lm(data=LitterCleaned, temperature_C~lakename) 
summary(LitterLakeName2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature. Both the lm and aov functions show that the p-value is < 2.2e-16. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.Scatter plot for temp, depth, with lakenames
LitterScatter2 <- ggplot(LitterCleaned, aes(x=depth, y=temperature_C, color=lakename))+
  geom_point(alpha=.5)+
  geom_smooth(method = "lm", se=FALSE)+
  ylim(0,35)+
  labs(x="Depth (m)", y="Temperature (C)", color="")+
  theme(legend.text = element_text(size=10))+
  guides(color=guide_legend(nrow=3, byrow=TRUE))


print(LitterScatter2)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 Tukey to find lake means

TukeyHSD(LitterLakeName)

LitterTukey<- HSD.test(LitterLakeName, "lakename", group=TRUE)
print(LitterTukey)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Lakes similar in mean temperatature to Peter Lake include Paul Lake and Ward Lake.  Central Long Lake and Crampton Lake are fairly distinct, with means of 17.66 and 15.35 respectively. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were looking at just Peter and Paul Lake, I would run a 2 Sample T-test to compare their means.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#T Test for two lakes
LitterCleaned2 <- LitterCleaned %>%
  filter(lakename== c("Ward Lake", "Crampton Lake"))

Litter2Sample <- t.test(LitterCleaned2$temperature_C~LitterCleaned2$lakename)
print(Litter2Sample)

```

>Answer: The p-value of the t test is .5484. Since this is more than .05, the mean temperatures are not significantly different. This matches the TukeyHSD findings in Question16. 
