---
title: "Assignment 2: Coding Basics"
author: Erin Ansbro
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. creating a sequence with from to by
firstseq <- seq(1,30,3) #naming sequence firstsequence
firstseq

#2. calculating mean and median of sequence
mean(firstseq)
median(firstseq)

#3. determining if median is greater than median
mean(firstseq) > median (firstseq)
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
Names <- c('Emma','John','Luana','Luke') #characters
Names

Grades <- c(81,99,99,49) #integer
Grades

Pass <- c(Grades >= 50) #logical
Pass

df_creation <- as.data.frame(Names) #creating data frame, changing vector to data frame
df_creation
df_NickSchoolClassGrades <- cbind(df_creation, Grades, Pass) #adding columns to data frame
df_NickSchoolClassGrades

#renaming columns to something informative
colnames(df_NickSchoolClassGrades) <- c('Names_of_Students', 'Grades_of_Students', 'Did_Student_Pass')
df_NickSchoolClassGrades 
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different because it has different types of data, e.g. numeric, logical, and characters. A matrix needs to have the same type of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
testvector <- c(20, 30, 50, 60)
testvector_passinggrades <- ifelse(testvector>=50, 'True', 'False')
print(testvector_passinggrades)

NickSchoolClassGrades_Passed <- ifelse(Grades>=50, 'True', 'False')
print(NickSchoolClassGrades_Passed)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The ifelse worked for me, I had trouble with 'if' and 'else' because it would throw up an error, for example:
x <- c(10, 40, 60, 80)
questionthree <- function(x) {
  if(x >= 50) {
    "True"
  } 
  else {
    "False"
  }
}
questionthree(x)
When I pressed run, Rstudio would print 'Error in if (x >= 50) { : the condition has length > 1'. It appears vectors are not allowed in an 'if' and 'else' statement only 'ifelse' statements, because they are more than a length of one element.
